---
title: "Seine & eDNA detection probability comparison"
author: "Lara Katz"
date: "2024-01-03"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries:
```{r}
library(emdbook)
library(ggplot2)
library(gridExtra)
library(R2WinBUGS) 
library(tidyverse)
```

**Install WinBUGS before continuing**

# Seine survey occupancy 

Bring in seine capture data:
```{r}
seine <- read.csv("./occu_seine.csv", header = TRUE)
seine$Waterbody <- as.factor(seine$Waterbody)
seine$Site.No <- as.factor(seine$Site.No)
```

Pivot seine haul values to create input csv files for the occupancy model. Each seine haul will have a separate csv:
```{r}
rep.cols <- c("s1", "s2", "s3", "s4", "s5", "s6", "s7", "s8")

for (i in 1:length(rep.cols)){
occu.seine <- seine %>% select(Waterbody, Site.No, rep.cols[i]) %>% 
  pivot_wider(., names_from = Site.No, values_from = rep.cols[i]) %>% 
  arrange(Waterbody)
colnames(occu.seine) <- c("","Site1","Site2","Site3","Site4")
write.csv(occu.seine, paste0(paste0("./seine_rep", i), "_detcompare.csv"), row.names = FALSE)
}
```

Import seine data (split by seine haul):
```{r}
haul1 <- read.csv("./seine_rep1_detcompare.csv", header=TRUE, row.names = 1) 
haul2 <- read.csv("./seine_rep2_detcompare.csv", header=TRUE, row.names = 1)  
haul3 <- read.csv("./seine_rep3_detcompare.csv", header=TRUE, row.names = 1)  
haul4 <- read.csv("./seine_rep4_detcompare.csv", header=TRUE, row.names = 1) 
haul5 <- read.csv("./seine_rep5_detcompare.csv", header=TRUE, row.names = 1)
haul6 <- read.csv("./seine_rep6_detcompare.csv", header=TRUE, row.names = 1) 
haul7 <- read.csv("./seine_rep7_detcompare.csv", header=TRUE, row.names = 1) 
haul8 <- read.csv("./seine_rep8_detcompare.csv", header=TRUE, row.names = 1) 
```

Create empty array of 8, 29x4 matrices:
```{r}
A <- array(as.numeric(NA), dim = c(29,4,8))
```

Fill array with seine data. Array dimensions are [1:29 (waterbodies), 1:4 (sites), 1:8 (seine hauls)]:
```{r}
A[,,1]<-as.matrix(haul1)
A[,,2]<-as.matrix(haul2) 
A[,,3]<-as.matrix(haul3) 
A[,,4]<-as.matrix(haul4) 
A[,,5]<-as.matrix(haul5)
A[,,6]<-as.matrix(haul6) 
A[,,7]<-as.matrix(haul7) 
A[,,8]<-as.matrix(haul8)

y <- A
```

## WinBUGS model

Format data for WinBUGS:
```{r}
str( win.data <- list(y = y, # data used to fill array
                      n.wb = dim(y)[1], # number of rows = number of waterbodies (29)
                      n.site = dim(y)[2], # number of columns = number of sites/waterbody (up to 4)
                      n.haul = dim(y)[3] )) # number of water sample replicates (up to 5)
```

Define model in BUGS language:
```{r}
sink("seine.model.txt") 

cat("
model{
# Priors and model for params 
  int.psi ~ dunif(0,1) # Intercept of occupancy probability (waterbodies) 
  for(t in 1:n.site){ 
      int.theta[t] ~ dunif(0,1)} # Intercepts of availability probability per site replicate
                             
  for(t in 1:n.haul){ 
      int.p[t] ~ dunif(0,1)}  # Intercepts of detection probability (1-replicate error)
      
# 'Likelihood' (or basic model structure)
  
  ## Occurrence in waterbody i 
  for(i in 1:n.wb){ 
      z[i] ~ dbern(psi[i]) 
      logit(psi[i]) <- logit(int.psi) 
      
      ## Occurrence in site j 
      for(j in 1:n.site){ 
          a[i,j] ~ dbern(mu.a[i,j])
          mu.a[i,j] <- z[i] * theta[i,j]
          logit(theta[i,j]) <- logit(int.theta[j])
  
          ## Detection error process in seine haul k 
          for (k in 1:n.haul){ 
               y[i,j,k] ~ dbern(mu.y[i,j,k]) 
               mu.y[i,j,k] <- a[i,j] * p[i,j,k] 
               logit(p[i,j,k]) <- logit(int.p[k])
                            }
                          }
                tmp[i] <- step(sum(a[i,])-0.1) 
                        }

# Derived quantities 
sum.z <- sum(z[]) # Total number of occupied waterbodies 
sum.a <- sum(tmp[]) # Total number of sites with presence 
mean.p <- mean(int.p[]) # mean p across seine haul replicates 
mean.theta <- mean(int.theta[]) # mean theta across site replicates 

} # end model 
",fill=TRUE)

sink()
```

Initial values:
```{r}
zst <- apply(y, 1, max) # inits for presence (z) 
ast <- apply(y, c(1,2), max) # inits for availability (a): applies the "max" function to all matrix rows and columns in the array
inits <- function() list(z = zst, a = ast, int.psi = 0.5)
```

Parameters:
```{r}
params <- c("int.psi", "int.theta", "int.p", "sum.z", "sum.a", "mean.p","mean.theta", "z") 
```

MCMC settings: 
```{r}
ni <- 25000 ; nt <- 10 ; nb <- 2000 ; nc <- 3
```

Call WinBUGS and summarize posterior:
```{r, cache=TRUE}
bd <- "C:/Program Files/winbugs14_full_patched/WinBUGS14" # Location of WinBUGS

seine.out <- bugs(win.data, inits, params, "seine.model.txt", 
                 n.chains = nc, n.thin = nt, 
                 n.iter = ni, n.burnin = nb, 
                 debug = TRUE, bugs.seed = 42,
                 bugs.dir = bd)

print(seine.out, 4)
```

```{r}
z.summary.seine <- as.data.frame(cbind(rownames(haul1), seine.out$mean$z, seine.out$sd$z))
colnames(z.summary.seine) <- c("Waterbody", "Psi.est", "sd")
z.summary.seine
min(z.summary.seine$Psi.est)
```

# Calculate p-star and theta-star (code from Erik Blomberg)

Calculate p*:
```{r}
hauls <- read.csv("./Number_samples_per_site.csv", header = TRUE)
mean.hauls <- mean(hauls$No.hauls)
sd.hauls <- sd(hauls$No.hauls)
mean.hauls - sd.hauls*1.96

no.hauls <- c(mean.hauls,seq(1,10,1)) # 1-10 seine hauls

pstar.seine <- 1-((1-seine.out$mean$mean.p)^no.hauls)

sigma.p.seine <- matrix(c(seine.out$sd$mean.p^2,0, 
                  0,0), nrow=2)

pstar.se.seine <- vector(length=length(pstar.seine)) 

for (i in 1:length(no.hauls)){
  pstar.se.seine[i]<- sqrt(deltavar(1-((1-x)^y),
                              meanval = c(x=seine.out$mean$mean.p, y=no.hauls[i]),
                              Sigma=sigma.p.seine)) }

pstar.df.seine <- data.frame(hauls=no.hauls, pstar = pstar.seine, se=pstar.se.seine)
```

95% Confidence Intervals:
```{r}
pstar.df.seine$lower <- pstar.df.seine$pstar - pstar.df.seine$se*1.96
pstar.df.seine$upper <- pstar.df.seine$pstar + pstar.df.seine$se*1.96
pstar.df.seine$upper[pstar.df.seine$upper > 1] <- 1
write.csv(pstar.df.seine, "./pstar_seine.csv")
pstar.df.seine
```

Pstar plot:
```{r}
pstar.95 <- 6 # number of seine hauls needed for detection to exceed 95%
pstar.80 <- 3 #80% 
pstar.50 <- 2 #50%

pstar.plot.seine <- ggplot(data=pstar.df.seine, aes(x=hauls, y=pstar))+
  geom_ribbon(aes(x=hauls,ymin=`lower`, ymax=`upper`),fill="gray", alpha=.6)+
  geom_line(linewidth=1)+
  geom_vline(xintercept=pstar.80, color='gray10', linetype="dashed", linewidth=0.8)+
  ylab("Cumulative detection probability (p*)")+
  xlab("Number of seine net hauls per site")+
  theme_classic()+
  theme(text = element_text(size = 20)) +
  coord_cartesian(ylim = c(0.1,1), xlim = c(1,10)) +
  annotate("text", x = Inf, y = Inf, label = paste("D"), vjust = 24, hjust = 1.5, size = 8)
pstar.plot.seine
```

Calculate theta*:
```{r}
sites.no <- read.csv("./Number_sites_per_waterbody.csv", header = TRUE)
mean.sites <- mean(sites.no$No.sites)

no.sites <- c(mean.sites, seq(1,10,1)) # Number of site replicates

tstar.seine <- 1-((1-seine.out$mean$mean.theta)^no.sites)

sigma.seine.t <- matrix(c(seine.out$sd$mean.theta^2,0, 
                  0,0), nrow=2)

tstar.se.seine <- vector(length=length(tstar.seine))

for (i in 1:length(no.sites)){
    tstar.se.seine[i]<- sqrt(deltavar(1-((1-x)^y),
                       meanval = c(x=seine.out$mean$mean.theta, y=no.sites[i]),
                       Sigma=sigma.seine.t)) }

tstar.df.seine <- data.frame(sites=no.sites, tstar = tstar.seine, se=tstar.se.seine)
```

95% confidence intervals:
```{r}
tstar.df.seine$lower <- tstar.df.seine$tstar - tstar.df.seine$se*1.96
tstar.df.seine$upper <- tstar.df.seine$tstar + tstar.df.seine$se*1.96
tstar.df.seine$upper[tstar.df.seine$upper > 1] <- 1
write.csv(tstar.df.seine, "./thetastar_seine.csv")
tstar.df.seine
```

Theta* plot:
```{r}
tstar.95 <- 5 # number of sites needed for detection to exceed 95%
tstar.80 <- 3 #80% 
tstar.50 <- 2 #50%

tstar.plot.seine<- ggplot(data=tstar.df.seine, aes(x=sites, y=tstar.seine))+
  geom_ribbon(aes(x=sites,ymin=`lower`, ymax=`upper`),fill="gray", alpha=.6)+
  geom_line(linewidth=1)+
  geom_vline(xintercept=tstar.80, color='gray10', linetype="dashed", linewidth=0.8)+
  ylab(expression("Cumulative availability probability (" ~ theta ~ "*)")) +
  xlab("Number of sites per waterbody (seine)")+
  theme_classic()+
  theme(text = element_text(size = 20)) +
  coord_cartesian(ylim = c(0.1,1), xlim = c(1,10)) +
  annotate("text", x = Inf, y = Inf, label = paste("C"), vjust = 24, hjust = 1.5, size = 8)
tstar.plot.seine
```

# eDNA survey occupancy (presence threshold: Cq = 50)

Bring in eDNA data:
```{r}
edna <- read.csv("./occu_edna_cq50.csv", header = TRUE)
edna$Replicate <- as.factor(edna$Replicate)
edna$Waterbody <- as.factor(edna$Waterbody)
edna$Site.No <- as.factor(edna$Site.No)
```

Pivot
```{r}
edna.wide <- edna %>% pivot_wider(., names_from = Replicate, values_from = all_cq)
edna.wide
```

Pivot water sample replicate values to create input csv files for the occupancy model. Each sample replicate will have a separate csv:
```{r}
rep.cols <- c("R1", "R2", "R3", "R4", "R5")

for (i in 1:length(rep.cols)){
occu.edna <- edna.wide %>% select(Waterbody, Site.No, rep.cols[i]) %>% 
  pivot_wider(., names_from = Site.No, values_from = rep.cols[i]) %>% 
  arrange(Waterbody)
colnames(occu.edna) <- c("","Site1","Site2","Site3","Site4")
write.csv(occu.edna, paste0(paste0("./edna_rep", i), "_detcompare_cq50.csv"), row.names = FALSE)
}
```

Import eDNA data (split by water sample replicate). Most samples have 2 replicates:
```{r}
rep1 <- read.csv("./edna_rep1_detcompare_cq50.csv", header=TRUE, row.names = 1) 
rep2 <- read.csv("./edna_rep2_detcompare_cq50.csv", header=TRUE, row.names = 1)  
rep3 <- read.csv("./edna_rep3_detcompare_cq50.csv", header=TRUE, row.names = 1)  
rep4 <- read.csv("./edna_rep4_detcompare_cq50.csv", header=TRUE, row.names = 1) 
rep5 <- read.csv("./edna_rep5_detcompare_cq50.csv", header=TRUE, row.names = 1) 
```

Create empty array of 5, 29x4 matrices:
```{r}
B <- array(as.numeric(NA), dim = c(29,4,5))
```

Fill array with eDNA data. Array dimensions are [1:29 (waterbodies), 1:4 (sites), 1:5 (water sample replicates)]:
```{r}
B[,,1]<-as.matrix(rep1)
B[,,2]<-as.matrix(rep2) 
B[,,3]<-as.matrix(rep3) 
B[,,4]<-as.matrix(rep4) 
B[,,5]<-as.matrix(rep5)
```

## WinBUGS model

Format data for WinBUGS:
```{r}
str( win.data <- list(y = B, # data used to fill array
                      n.wb = dim(B)[1], # number of rows = number of waterbodies (29)
                      n.site = dim(B)[2], # number of columns = number of sites/waterbody (up to 4)
                      n.rep = dim(B)[3] )) # number of water sample replicates (up to 5)
```

Define model in BUGS language:
```{r}
sink("eDNA.model.txt") 

cat("
model{
# Priors and model for params 
  int.psi ~ dunif(0,1) # Intercept of occupancy probability (waterbodies) 
  for(t in 1:n.site){ 
      int.theta[t] ~ dunif(0,1)} # Intercepts of availability probability per site replicate
                             
  for(t in 1:n.rep){ 
      int.p[t] ~ dunif(0,1)}  # Intercepts of detection probability (1-replicate error)
      
# 'Likelihood' (or basic model structure)
  
  ## Occurrence in waterbody i 
  for(i in 1:n.wb){ 
      z[i] ~ dbern(psi[i])
      logit(psi[i]) <- logit(int.psi) 
      
      ## Occurrence in site j 
      for(j in 1:n.site){ 
          a[i,j] ~ dbern(mu.a[i,j])
          mu.a[i,j] <- z[i] * theta[i,j]
          logit(theta[i,j]) <- logit(int.theta[j])
  
          ## Detection error process in replicate k 
          for (k in 1:n.rep){ 
               y[i,j,k] ~ dbern(mu.y[i,j,k]) 
               mu.y[i,j,k] <- a[i,j] * p[i,j,k] 
               logit(p[i,j,k]) <- logit(int.p[k])
                            }
                          }
                tmp[i] <- step(sum(a[i,])-0.1) 
                        }

# Derived quantities 
sum.z <- sum(z[]) # Total number of occupied waterbodies 
sum.a <- sum(tmp[]) # Total number of sites with presence 
mean.p <- mean(int.p[]) # mean p across water sample replicates 
mean.theta <- mean(int.theta[]) # mean theta across site replicates 

} # end model 
",fill=TRUE)

sink()
```

Initial values:
```{r}
zst <- apply(B, 1, max) # inits for presence (z) 
ast <- apply(B, c(1,2), max) # inits for availability (a): applies the "max" function to all matrix rows and columns in the array
inits <- function() list(z = zst, a = ast, int.psi = 0.5)
```

Parameters:
```{r}
params <- c("int.psi", "int.theta", "int.p", "sum.z", "sum.a", "mean.p","mean.theta", "z") 
```

MCMC settings: 
```{r}
ni <- 25000 ; nt <- 10 ; nb <- 2000 ; nc <- 3
```

Call WinBUGS and summarize posterior:
```{r, cache=TRUE}
bd <- "C:/Program Files/winbugs14_full_patched/WinBUGS14" # Location of WinBUGS

eDNA.out <- bugs(win.data, inits, params, "eDNA.model.txt", 
                 n.chains = nc, n.thin = nt, 
                 n.iter = ni, n.burnin = nb, 
                 debug = TRUE, bugs.seed = 42,
                 bugs.dir = bd)

print(eDNA.out, 4)
```

```{r}
z.summary.edna <- as.data.frame(cbind(rownames(rep1), eDNA.out$mean$z, eDNA.out$sd$z))
colnames(z.summary.edna) <- c("Waterbody", "Psi.est", "sd")
z.summary.edna
```

# Calculate p-star and theta-star (code from Erik Blomberg)

Calculate p*:
```{r}
samples <- read.csv("./Number_samples_per_site.csv", header = TRUE)
mean.samples <- mean(samples$No.eDNA)
sd.samples <- sd(samples$No.eDNA)
mean.samples - sd.samples*1.96


no.reps <- c(mean.samples, seq(1,10,1)) # 1-10 water sample replicates

pstar.edna <- 1-((1-eDNA.out$mean$mean.p)^no.reps)

sigma.p.edna <- matrix(c(eDNA.out$sd$mean.p^2,0, 
                  0,0), nrow=2)

pstar.se.edna <- vector(length=length(pstar.edna)) 

for (i in 1:length(no.reps)){
  pstar.se.edna[i]<- sqrt(deltavar(1-((1-x)^y),
                              meanval = c(x=eDNA.out$mean$mean.p, y=no.reps[i]),
                              Sigma=sigma.p.edna)) }

pstar.df.edna <- data.frame(rep=no.reps, pstar = pstar.edna, se=pstar.se.edna)
```

95% Confidence Intervals:
```{r}
pstar.df.edna$lower <- pstar.df.edna$pstar - pstar.df.edna$se*1.96
pstar.df.edna$upper <- pstar.df.edna$pstar + pstar.df.edna$se*1.96
pstar.df.edna$upper[pstar.df.edna$upper > 1] <- 1
write.csv(pstar.df.edna, "./pstar_edna_cq50.csv")
pstar.df.edna
```

Pstar plot:
```{r}
pstar.95 <- 5 # number of samples needed for detection to exceed 95%
pstar.80 <- 3 #80% 
pstar.50 <- 2 #50%

pstar.plot.edna <- ggplot(data=pstar.df.edna, aes(x=rep, y=pstar))+
  geom_ribbon(aes(x=rep,ymin=`lower`, ymax=`upper`),fill="gray", alpha=.6)+
  geom_line(linewidth=1)+
  geom_vline(xintercept=pstar.80, color='gray10', linetype="dashed", linewidth=0.8)+
  ylab("Cumulative detection probability (p*)")+
  xlab("Number of 1-L eDNA samples per site")+
  theme_classic()+
  theme(text = element_text(size = 20)) +
  coord_cartesian(ylim = c(0.1,1), xlim = c(1,10)) +
  annotate("text", x = Inf, y = Inf, label = paste("B"), vjust = 24, hjust = 1.5, size = 8)
pstar.plot.edna
```

Calculate theta*:
```{r}
sites.no <- read.csv("./Number_sites_per_waterbody.csv", header = TRUE)
mean.sites <- mean(sites.no$No.sites)
se.sites <- sd(sites.no$No.sites)
mean.sites + se.sites*1.96


no.sites <- c(mean.sites, seq(1,10,1)) # Number of site replicates

tstar.edna <- 1-((1-eDNA.out$mean$mean.theta)^no.sites)

sigma.edna.t <- matrix(c(eDNA.out$sd$mean.theta^2,0, 
                  0,0), nrow=2)

tstar.se.edna <- vector(length=length(tstar.edna))

for (i in 1:length(no.sites)){
    tstar.se.edna[i]<- sqrt(deltavar(1-((1-x)^y),
                       meanval = c(x=eDNA.out$mean$mean.theta, y=no.sites[i]),
                       Sigma=sigma.edna.t)) }

tstar.df.edna <- data.frame(sites=no.sites, tstar = tstar.edna, se=tstar.se.edna)
```

95% confidence intervals:
```{r}
tstar.df.edna$lower <- tstar.df.edna$tstar - tstar.df.edna$se*1.96
tstar.df.edna$upper <- tstar.df.edna$tstar + tstar.df.edna$se*1.96
tstar.df.edna$upper[tstar.df.edna$upper > 1] <- 1
write.csv(tstar.df.edna, "./thetastar_edna_cq50.csv")
tstar.df.edna
```

Theta* plot:
```{r}
tstar.95 <- 5 # number of samples needed for detection to exceed 95%
tstar.80 <- 3 #80% 
tstar.50 <- 1 #50%

tstar.plot.edna<- ggplot(data=tstar.df.edna, aes(x=sites, y=tstar.edna))+
  geom_ribbon(aes(x=sites,ymin=`lower`, ymax=`upper`),fill="gray", alpha=.6)+
  geom_line(linewidth=1)+
  geom_vline(xintercept=tstar.80, color='gray10', linetype="dashed", linewidth=0.8)+
  ylab(expression("Cumulative availability probability (" ~ theta ~ "*)")) +
  xlab("Number of sites per waterbody (eDNA)")+
  theme_classic()+
  theme(text = element_text(size = 20)) +
  coord_cartesian(ylim = c(0.1,1), xlim = c(1,10)) +
  annotate("text", x = Inf, y = Inf, label = paste("A"), vjust = 24, hjust = 1.5, size = 8)
tstar.plot.edna
```

Save plots:
```{r, }
#png(filename="./detcompare.png", width = 1200, height = 1000)
ggsave <- grid.arrange(tstar.plot.edna, pstar.plot.edna, tstar.plot.seine, pstar.plot.seine, nrow = 2)
ggsave(filename = "./detcompare.png", plot = ggsave, width = 12.6, height = 12.6, dpi = 600)
#dev.off()
```