---
title: "Supplement B. Community composition"
author: "Lara Katz"
date: "2024-01-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(vegan)
library(pracma)
library(ggplot2)
library(dplyr)
library(gridExtra)
```

Read in species matrix and grouping variables:
```{r}
comm <- read.csv("./CPUE_2021_bySite.csv", header = TRUE, row.names = 1)
```

Remove waterbody with no fish captured (GWORKN):
```{r}
comm <- comm[-8,]
rownames(comm)
```

Read in grouping data:
```{r}
comm_groups <- read.csv("./nmds_groups.csv", row.names = 1)
comm_groups$BDS.pres <- as.factor(comm_groups$BDS.pres) # Bridle Shiner presence
# comm_groups$Size <- as.factor(comm_groups$Size) # waterbody size/catchment position
```

Fourth-root transformation of catch per unit effort (CPUE) values:
```{r}
df <- data.frame(matrix(ncol = 26, nrow = 28))
colnames(df) <- colnames(comm)
row.names(df) <- rownames(comm)

for (i in 1:ncol(comm)){
  df[,i] <- nthroot(comm[,i], 4)
}
```

Calculate distance matrix and save to csv:
```{r}
comm.distmat <- as.matrix(vegdist(df, method = "bray"), labels = TRUE)
write.csv(comm.distmat, "./community_distmat.csv")
```

Run NMDS in `vegan` using the `metaMDS` function:
```{r}
set.seed(42)
comm_NMDS <- metaMDS(df, distance = "bray", 
                     k = 3, maxit = 999, 
                     trymax = 500, 
                     wascores = TRUE,
                     tidy = TRUE)
```

As a rule of thumb literature has identified the following cut-off values for stress-level:

-   Higher than 0.2 is poor (risks for false interpretation).
-   0.1 - 0.2 is fair (some distances can be misleading for interpretation).
-   0.05 - 0.1 is good (can be confident in inferences from plot).
-   Less than 0.05 is excellent (this can be rare).

Save species scores:
```{r}
scores <- as.data.frame(comm_NMDS$species)
```

Shepards test/goodness of fit:
```{r}
goodness(comm_NMDS) # Produces a result of test statistics for goodness of fit for each point

stressplot(comm_NMDS) # Produces a Shepards diagram
```

Plotting points in ordination space:
```{r}
plot(comm_NMDS, "sites")   # Produces distance 
orditorp(comm_NMDS, "sites")   # Label points
```

Plot waterbodies and species along NMDS axes:
```{r, fig.height=5, fig.width=12}
#png("./NMDSplots.png", width = 1050)
par(mfrow = c(1,2)) 
p1 <- ordiplot(ord = comm_NMDS, 
               choices = c(1,2), 
               type = "none", 
               display = c("sites", "species"))
text(p1, "sites", col = "black", cex = 1)
ordiellipse(comm_NMDS, comm_groups$BDS.pres, 
            display = "sites", 
            draw = "polygon", 
            label = F, kind = "ehull")
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)
mtext("A", side = 1, adj = 1, line = -1.5, at = 2.4, cex = 2)

p2 <- ordiplot(ord = comm_NMDS, 
               choices = c(1,2), 
               type = "none", 
               display = c("sites", "species"))
text(p2, "species", col = "black", cex = 1)
ordiellipse(comm_NMDS, comm_groups$BDS.pres,
            display = "sites", 
            draw = "polygon", 
            label = F, kind = "ehull")
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)
mtext("B", side = 1, adj = 1, line = -1.5, at = 2.4, cex = 2)
#dev.off()
```

Use `envfit` to determine which species were the intrinsic drivers of the waterbody distribution pattern:
```{r}
spp.fit <- envfit(comm_NMDS, df, permutations = 999)
head(spp.fit)
```

Plot intrinsic species:
```{r}
#png("./IntrinsicSpp.png")
ordiplot(comm_NMDS, type = "n")
orditorp(comm_NMDS, display = "sites", labels = T,
         #pch = c(1,16)[comm_groups$BDS.pres], 
         cex = 1)
plot(spp.fit, 
     p.max = 0.05, # change the significance level of species shown
     col = "black", 
     cex = 1) 
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)
#dev.off()
```
