---
title: "Supplement B. Community composition"
author: "Lara Katz"
date: "2024-01-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(vegan)
library(pracma)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(scatterplot3d)
library(vegan3d)
library(labdsv)
```

Read in species matrix and grouping variables:
```{r}
comm <- read.csv("./CPUE_2021_bySite.csv", header = TRUE, row.names = 1)
```

Remove waterbody with no fish captured (GWORKN):
```{r}
comm <- comm[-8,]
rownames(comm)
```

Read in grouping data:
```{r}
comm_groups <- read.csv("./nmds_groups.csv", row.names = 1)
comm_groups$BDS.pres <- as.factor(comm_groups$BDS.pres) # Bridle Shiner presence
# comm_groups$Size <- as.factor(comm_groups$Size) # waterbody size/catchment position
```

Fourth-root transformation of catch per unit effort (CPUE) values:
```{r}
df <- data.frame(matrix(ncol = 26, nrow = 28))
colnames(df) <- colnames(comm)
row.names(df) <- rownames(comm)

for (i in 1:ncol(comm)){
  df[,i] <- nthroot(comm[,i], 4)
}
```

Calculate distance matrix and save to csv:
```{r}
comm.distmat <- as.matrix(vegdist(df, method = "bray"), labels = TRUE)
write.csv(comm.distmat, "./community_distmat.csv")
```

# NMDS with two axes

```{r}
set.seed(42)
comm_NMDS <- metaMDS(df, distance = "bray", 
                     k = 2, try = 500, trymax = 600,
                     wascores = TRUE,
                     tidy = TRUE)
comm_NMDS
```

Shepards test/goodness of fit:
```{r}
jpeg("./ShepardPlot.jpg", width = 900, height = 600, res = 100)
stressplot(comm_NMDS) # Produces a Shepards diagram
dev.off()
```

Plot waterbodies and species:
```{r, fig.height=5, fig.width=12}
#png("./NMDSplots.png", width = 1000)
par(mfrow = c(1,2))
p1 <- ordiplot(ord = comm_NMDS,
               choices = c(1,2),
               type = "none",
               display = c("sites", "species"))
text(p1, "sites", col = "black")
ordiellipse(comm_NMDS, comm_groups$BDS.pres,
            display = "sites",
            draw = "polygon",
            col = c("red", "blue"),
            label = F, kind = "se", alpha = 80, conf = 0.95)
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)
mtext("A", side = 1, adj = 1, line = -1.5, at = 1.8, cex = 2)

p2 <- ordiplot(ord = comm_NMDS,
               choices = c(1,2),
               type = "none",
               display = c("sites", "species"))
text(p2, "species", col = "black")
ordiellipse(comm_NMDS, comm_groups$BDS.pres,
            display = "sites",
            draw = "polygon",
            col = c("red", "blue"),
            label = F, kind = "se", alpha = 80, conf = 0.95)
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)
mtext("B", side = 1, adj = 1, line = -1.5, at = 1.8, cex = 2)
#dev.off()
```

## Envfit

See https://jkzorz.github.io/2020/04/04/NMDS-extras.html or https://www.rpubs.com/RGrieger/545184 for more info on `envfit`.

Use `envfit` to determine which species were the intrinsic drivers of the waterbody distribution pattern:
```{r}
spp.fit <- envfit(comm_NMDS, df, permutations = 999)
head(spp.fit)
```

Scale arrow segments to R2 value and to the plot:
```{r}
env_coord <- as.data.frame(vegan::scores(spp.fit, display = "vectors")) * ordiArrowMul(spp.fit)
env_coord[order(env_coord$NMDS1),]
```

Extract manually and limit to species with p-values < 0.05:
```{r}
coords <- as.data.frame(cbind(spp.fit$vectors$arrows, spp.fit$vectors$r, spp.fit$vectors$pvals))
coords[order(coords$V4),]

coords.05 <- coords[coords$V4 < 0.05,]
coords.05$NMDS1.scale <- coords.05$NMDS1 * sqrt(coords.05$V3) * ordiArrowMul(spp.fit)
coords.05$NMDS2.scale <- coords.05$NMDS2 * sqrt(coords.05$V3) * ordiArrowMul(spp.fit)
coords.05
```

Which species influenced the NMDS1 axis the most?
```{r}
rownames(coords.05[which(coords.05$NMDS1.scale == max(coords.05$NMDS1.scale)),])
max(coords.05$NMDS1.scale) # most + influential
rownames(coords.05[which(coords.05$NMDS1.scale == min(coords.05$NMDS1.scale)),])
min(coords.05$NMDS1.scale) # most - influential
```

Which species influenced the NMDS1 axis the most?
```{r}
rownames(coords.05[which(coords.05$NMDS2.scale == max(coords.05$NMDS2.scale)),])
max(coords.05$NMDS2.scale) # most + influential
rownames(coords.05[which(coords.05$NMDS2.scale == min(coords.05$NMDS2.scale)),])
min(coords.05$NMDS2.scale) # most - influential
```

Plot `envfit` results:
```{r}
jpeg("./IntrinsicSpp2d.jpg", width = 600, height = 600, res = 100)
ordiplot(comm_NMDS, type = "n")
# orditorp(comm_NMDS, display = "sites", labels = T,
#          #pch = c(1,16)[comm_groups$BDS.pres], 
#          cex = 1)
plot(spp.fit, 
     p.max = 0.05, # change the significance level of species shown
     col = "black", 
     cex = 1) 
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)
dev.off()
```

# NMDS with three axes

Run NMDS in `vegan` using the `metaMDS` function:
```{r}
set.seed(42)
comm_NMDS <- metaMDS(df, distance = "bray", 
                     k = 3, try = 500, trymax = 600,
                     wascores = TRUE,
                     tidy = TRUE)
comm_NMDS
```

As a rule of thumb literature has identified the following cut-off values for stress-level:

-   Higher than 0.2 is poor (risks for false interpretation).
-   0.1 - 0.2 is fair (some distances can be misleading for interpretation).
-   0.05 - 0.1 is good (can be confident in inferences from plot).
-   Less than 0.05 is excellent (this can be rare).

Save species scores:
```{r}
scores <- as.data.frame(comm_NMDS$species)
```

Shepards test/goodness of fit:
```{r}
#jpeg("./ShepardPlot.jpg", width = 900, height = 600, res = 100)
par(mfrow= c(1,2))
stressplot(comm_NMDS) # Produces a Shepards diagram
#dev.off()
```
Plots of sites (2D, all axes):
```{r}
gof <- goodness(comm_NMDS) # Produces a result of test statistics for goodness of fit for each point

jpeg("./nmds2dsites.jpg", width = 2100, height = 600, res = 200)

par(mfrow=c(1,3))
p1 <- plot(comm_NMDS, type = "n", choices = c(1,2))
text(p1, "sites", col = "black")
ordiellipse(comm_NMDS, comm_groups$BDS.pres,
            choices = c(1,2),
            display = "sites",
            col = c("red", "blue"),
            draw = "polygon", 
            label = F, kind = "sd")
#points(comm_NMDS, display = "sites", cex = 2*gof/mean(gof))
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)

p2 <- plot(comm_NMDS, type = "n", choices = c(1,3))
text(p2, "sites", col = "black")
ordiellipse(comm_NMDS, comm_groups$BDS.pres,
            choices = c(1,3),
            display = "sites",
            col = c("red", "blue"),
            draw = "polygon", 
            label = F, kind = "sd")
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)
#points(comm_NMDS, display = "sites", cex = 2*gof/mean(gof))

p3 <- plot(comm_NMDS, type = "n", choices = c(2,3))
text(p3, "sites", col = "black")
#points(comm_NMDS, display = "sites", cex = 2*gof/mean(gof))
ordiellipse(comm_NMDS, comm_groups$BDS.pres,
            choices = c(2,3),
            display = "sites", 
            col = c("red", "blue"),
            draw = "polygon", 
            label = F, kind = "sd")
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)
dev.off()
```
Plots of species (2D, all axes):
```{r}
jpeg("./nmds2dspecies.jpg", width = 2100, height = 600, res = 200)

par(mfrow=c(1,3))
p1 <- plot(comm_NMDS, type = "n", choices = c(1,2), ylim = c(-1,2))
text(p1, "species", col = "black")
ordiellipse(comm_NMDS, comm_groups$BDS.pres,
            choices = c(1,2),
            display = "sites",
            col = c("red", "blue"),
            draw = "polygon", 
            label = F, kind = "sd")
#points(comm_NMDS, display = "sites", cex = 2*gof/mean(gof))
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)

p2 <- plot(comm_NMDS, type = "n", choices = c(1,3))
text(p2, "species", col = "black")
ordiellipse(comm_NMDS, comm_groups$BDS.pres,
            choices = c(1,3),
            display = "sites",
            col = c("red", "blue"),
            draw = "polygon", 
            label = F, kind = "sd")
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)
#points(comm_NMDS, display = "sites", cex = 2*gof/mean(gof))

p3 <- plot(comm_NMDS, type = "n", choices = c(2,3))
text(p3, "species", col = "black")
#points(comm_NMDS, display = "sites", cex = 2*gof/mean(gof))
ordiellipse(comm_NMDS, comm_groups$BDS.pres,
            choices = c(2,3),
            display = "sites", 
            col = c("red", "blue"),
            draw = "polygon", 
            label = F, kind = "sd")
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)
dev.off()
```
3D plots:
```{r}
p1 <- ordirgl(comm_NMDS, type = "t")
sites <- scores(comm_NMDS, choices=1:3, display="sites")
#text(p1$xyz.convert(sites), rownames(sites), cex=0.7, xpd=TRUE)
orgltext(comm_NMDS, display = "sites")
with(comm_groups, orglellipse(comm_NMDS, comm_groups$BDS.pres, col = c("red", "blue"), kind = "sd"))
# points(pl, "points", pch=16, col="black", cex = 0.7)


p2 <- ordirgl(comm_NMDS, type = "t", display = "species")
spp <- scores(comm_NMDS, choices=1:3, display="species")
orgltext(comm_NMDS, display = "species")
with(comm_groups, orglellipse(comm_NMDS, comm_groups$BDS.pres, col = c("red", "blue"), kind = "sd"))
```

## Envfit

See https://jkzorz.github.io/2020/04/04/NMDS-extras.html or https://www.rpubs.com/RGrieger/545184 for more info on `envfit`.

Use `envfit` to determine which species were the intrinsic drivers of the waterbody distribution pattern:
```{r}
spp.fit <- envfit(comm_NMDS, df, permutations = 999, choices = 1:3)
head(spp.fit)
env_coord <- as.data.frame(vegan::scores(spp.fit, "vectors"))* ordiArrowMul(spp.fit)
```

Plot intrinsic species with a flattened 3D plot:
```{r}
#jpeg("./IntrinsicSpp3d.jpg", width = 1000, height = 600, res = 100)

p4 <- ordiplot3d(comm_NMDS, type = "n", box = FALSE, angle = 160, scale.y = 0.5)
plot(p4$envfit.convert(spp.fit), at = p4$origin, 
     p.max = 0.05, # change the significance level of species shown
     col = "black", 
     cex = 1) 

#dev.off()

```
Which species influenced the NMDS1 axis the most?
```{r}
rownames(env_coord[which(env_coord[,1] == max(env_coord[,1])),])
max(env_coord[,1]) # most + influential
rownames(env_coord[which(env_coord[,1] == min(env_coord[,1])),])
min(env_coord[,1]) # most - influential
```

Which species influenced the NMDS2 axis the most?
```{r}
rownames(env_coord[which(env_coord[,2] == max(env_coord[,2])),])
max(env_coord[,2]) # most + influential
rownames(env_coord[which(env_coord[,2] == min(env_coord[,2])),])
min(env_coord[,2]) # most - influential
```

Which species influenced the NMDS3 axis the most?
```{r}
rownames(env_coord[which(env_coord[,3] == max(env_coord[,3])),])
max(env_coord[,3]) # most + influential
rownames(env_coord[which(env_coord[,3] == min(env_coord[,3])),])
min(env_coord[,3]) # most - influential
```

