---
title: "Prepare data for WinBUGS"
author: "Lara Katz"
date: "2023-12-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries:
```{r}
library(tidyverse)
library(lubridate)
```

Set Cq threshold value:
```{r}
cq <- 50
```

Import data:
```{r}
data <- read.csv("./eDNA_results_2021-2023.csv", header = TRUE)

data1 <- transform(data,
                   Filter.Number = Filter.Number,
                   Filtering.Date = mdy(Filtering.Date),
                   Site = as.factor(Site),
                   Replicate = Replicate,
                   Volume.Filtered..mL. = as.integer(Volume.Filtered..mL.),
                   Combined.Sample. = as.factor(Combined.Sample.),
                   qpcr.date = as.factor(qpcr.date),
                   pos = as.factor(pos),
                   no.reps.pos = as.character(no.reps.pos),
                   cq1 = as.numeric(cq1),
                   cq2 = as.numeric(cq2),
                   cq3 = as.numeric(cq3),
                   cq4 = as.numeric(cq4),
                   cq5 = as.numeric(cq5),
                   cq6 = as.numeric(cq6),
                   cq7 = as.numeric(cq7),
                   cq8 = as.numeric(cq8),
                   notes = notes,
                   plate = as.factor(plate))
```

Filter out field controls and samples with no PCR results:
```{r}
rm <- list("Control", "B-R1", "B-R2", "B-R3", "B-R4", "B-R5",
        "C-R1", "C-R2", "C-R3", "C-R4", "C-R5")
data.pcr <- data1[!is.na(data1$pos),] %>%  # remove NAs (site HALEY)
  filter(., Volume.Filtered..mL. > 100) # remove LITTLP (84mL)
  
for (i in 1:length(rm)) {
  data.pcr <- filter(data.pcr, Replicate != rm[[i]])
}

data.pcr[data.pcr$Replicate == "A-R1",]$Replicate <- "R1"
data.pcr[data.pcr$Replicate == "A-R2",]$Replicate <- "R2"
data.pcr[data.pcr$Replicate == "A-R3",]$Replicate <- "R3"
data.pcr[data.pcr$Replicate == "A-R4",]$Replicate <- "R4"
data.pcr[data.pcr$Replicate == "A-R5",]$Replicate <- "R5"

summary(data.pcr)
```

## All historic and new sites

Select only the first sampling occasion for sites sampled more than once:
```{r}
ossipe.min <- min(data.pcr[data.pcr$Site == "OSSIPE",]$Filtering.Date) # seasonal site
ossipe <- data.pcr[data.pcr$Site == "OSSIPE" & data.pcr$Filtering.Date == ossipe.min,]

presum.min <- min(data.pcr[data.pcr$Site == "PRESUM-01",]$Filtering.Date) # seasonal site
presum <- data.pcr[data.pcr$Site == "PRESUM-01" & data.pcr$Filtering.Date == presum.min,]

watbrk.min <- min(data.pcr[data.pcr$Site == "WATBRK",]$Filtering.Date) # seasonal site
watbrk <- data.pcr[data.pcr$Site == "WATBRK" & data.pcr$Filtering.Date == watbrk.min,]

barker.min <- min(data.pcr[data.pcr$Site == "BARKER",]$Filtering.Date) # filter pore size site
barker <- data.pcr[data.pcr$Site == "BARKER" & data.pcr$Filtering.Date == barker.min,]

data.pcr2 <- data.pcr %>% filter(., Site != "BARKER") %>% filter(., Site != "PRESUM-01") %>% filter(., Site != "OSSIPE") %>% filter(., Site != "WATBRK")

occu.sites <- rbind(data.pcr2, barker, presum, ossipe, watbrk)
```

Determine which Cq reps are positive (depending on the Cq threshold set above). Replicates that did not amplify are denoted by the 999 value:
```{r}
rep.cols <- c("cq1", "cq2", "cq3", "cq4", "cq5", "cq6", "cq7", "cq8")

occu.pres <- occu.sites %>% mutate(across(rep.cols, ~if_else(. <= cq, 1, 0), .names = "pres_{.col}"))
write.csv(occu.pres, "./occu_pres.csv", row.names = FALSE)
```

Pivot Cq values to create input csv files for the occupancy model. Each Cq replicate will have a separate csv:
```{r}
occu.pres$Replicate <- as.factor(occu.pres$Replicate)
rep.cols2 <- c("pres_cq1", "pres_cq2", "pres_cq3", "pres_cq4", "pres_cq5", "pres_cq6", "pres_cq7", "pres_cq8")

for (i in 1:length(rep.cols2)){
occu.pres2 <- occu.pres %>% select(Site, Replicate, rep.cols2[i]) %>% 
  pivot_wider(., names_from = Replicate, values_from = rep.cols2[i]) %>% 
  arrange(Site)
colnames(occu.pres2) <- c("","R1","R2","R3","R4","R5")
write.csv(occu.pres2, paste0(paste0("./qpcr_rep", i), "_cq50.csv"), row.names = FALSE) # change ouput file name depending on which threshold is used
}
```