---
title: "Bridle shiner 3-tiered hierarchical occupancy model in WinBUGS"
author: "Lara Katz adapted from code by Geneva York & Erik Blomberg"
date: "2024-01-03"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code is based on the hierarchical occupancy model described in:
Kéry, M., and Royle, J.A. 2016. Applied hierarchical modeling in ecology: analysis of distribution, abundance and species richness in R and BUGS. Academic Press/Elsevier, Amsterdam; Boston.

**Install WinBUGS before continuing.**

Load required packages:
```{r, warning=FALSE, message=FALSE}
library(emdbook)
library(ggplot2)
library(gridExtra)
library(R2WinBUGS)
```

# Presence threshold: Cq = 40

Import eDNA data (split by qPCR replicate). Most samples have 4 replicates but some have 8:
```{r}
pcr1 <- read.csv("./qpcr_rep1_cq40.csv", header=TRUE, row.names = 1) 
pcr2 <- read.csv("./qpcr_rep2_cq40.csv", header=TRUE, row.names = 1)  
pcr3 <- read.csv("./qpcr_rep3_cq40.csv", header=TRUE, row.names = 1)  
pcr4 <- read.csv("./qpcr_rep4_cq40.csv", header=TRUE, row.names = 1) 
pcr5 <- read.csv("./qpcr_rep5_cq40.csv", header=TRUE, row.names = 1) 
pcr6 <- read.csv("./qpcr_rep6_cq40.csv", header=TRUE, row.names = 1)  
pcr7 <- read.csv("./qpcr_rep7_cq40.csv", header=TRUE, row.names = 1)  
pcr8 <- read.csv("./qpcr_rep8_cq40.csv", header=TRUE, row.names = 1) 
```

Create empty array of 8, 93x5 matrices:
```{r}
A <- array(as.numeric(NA), dim = c(93,5,8))
```

Fill array with eDNA data. Array dimensions are [1:93 (sites), 1:5 (sample replicates), 1:8 (qPCR replicates)]:
```{r}
A[,,1]<-as.matrix(pcr1)
A[,,2]<-as.matrix(pcr2) 
A[,,3]<-as.matrix(pcr3) 
A[,,4]<-as.matrix(pcr4) 
A[,,5]<-as.matrix(pcr5)
A[,,6]<-as.matrix(pcr6) 
A[,,7]<-as.matrix(pcr7) 
A[,,8]<-as.matrix(pcr8) 

y <- A 
```

## WinBUGS model

Format data for WinBUGS:
```{r}
str( win.data40 <- list(y = y, # data used to fill array
                      n.site = dim(y)[1], # number of rows = number of sites (93)
                      n.samples = dim(y)[2], # number of columns = number of 1L replicate samples/site (up to 5)
                      n.pcr = dim(y)[3] )) # number of PCR replicates (up to 8)
```

Define model in BUGS language:
```{r, include=TRUE, eval=FALSE}
sink("eDNA.model40.txt") 

cat("
model{
# Priors and model for params 
  int.psi ~ dunif(0,1) # Intercept of occupancy probability (sites) 
  for(t in 1:n.samples){ 
      int.theta[t] ~ dunif(0,1)} # Intercepts of availability probability per sample replicate
                             
  for(t in 1:n.pcr){ 
      int.p[t] ~ dunif(0,1)}  # Intercepts of detection probability (1-PCR error)
      
# 'Likelihood' (or basic model structure)
  
  ## Occurrence in site i 
  for(i in 1:n.site){ 
      z[i] ~ dbern(psi[i]) 
      logit(psi[i]) <- logit(int.psi) 
      
      ## Occurrence in sample j 
      for(j in 1:n.samples){ 
          a[i,j] ~ dbern(mu.a[i,j])
          mu.a[i,j] <- z[i] * theta[i,j]
          logit(theta[i,j]) <- logit(int.theta[j])
  
          ## PCR detection error process in sample k 
          for (k in 1:n.pcr){ 
               y[i,j,k] ~ dbern(mu.y[i,j,k]) 
               mu.y[i,j,k] <- a[i,j] * p[i,j,k] 
               logit(p[i,j,k]) <- logit(int.p[k])
                            }
                          }
                tmp[i] <- step(sum(a[i,])-0.1) 
                        }

# Derived quantities 
sum.z <- sum(z[]) # Total number of occupied sites 
sum.a <- sum(tmp[]) # Total number of samples with presence 
mean.p <- mean(int.p[]) # mean p across qPCR replicates 
mean.theta <- mean(int.theta[]) # mean theta across sample replicates 

} # end model 
",fill=TRUE)

sink()
```

Initial values:
```{r}
zst <- apply(y, 1, max) # inits for presence (z) 
ast <- apply(y, c(1,2), max) # inits for availability (a): applies the "max" function to all matrix rows and columns in the array
inits <- function() list(z = zst, a = ast, int.psi = 0.5)
```

Parameters:
```{r}
params <- c("int.psi","int.theta","int.p","sum.z","sum.a","mean.p","mean.theta") 
```

MCMC settings:
```{r}
ni <- 25000 ; nt <- 10 ; nb <- 2000 ; nc <- 3
```

Call WinBUGS and summarize posterior:
```{r, include=TRUE, eval=FALSE}
bd <- "C:/Program Files/winbugs14_full_patched/WinBUGS14" # Location of WinBUGS

eDNA.out40 <- bugs(win.data40, inits, params, "eDNA.model40.txt", 
                 n.chains = nc, n.thin = nt, 
                 n.iter = ni, n.burnin = nb, 
                 debug = TRUE, bugs.seed = 42,
                 bugs.dir = bd)
saveRDS(eDNA.out40, "./eDNA_out40.RDS")
```

View posterior summary:
```{r, cache=TRUE}
eDNA.out40 <- readRDS("./eDNA_out40.RDS")
print(eDNA.out40, 4)
```

## Calculate p-star and theta-star (code from Erik Blomberg)

Calculate p*:
```{r}
no.pcr <- seq(1,10,1) # 1-10 qPCR replicates

pstar <- 1-((1-eDNA.out40$mean$mean.p)^no.pcr)

sigma <- matrix(c(eDNA.out40$sd$mean.p^2,0, 
                  0,0), nrow=2)

pstar.se <- vector(length=length(pstar)) 

for (i in 1:length(no.pcr)){
  pstar.se[i]<- sqrt(deltavar(1-((1-x)^y),
                              meanval = c(x=eDNA.out40$mean$mean.p, y=no.pcr[i]),
                              Sigma=sigma)) }

pstar.df <- data.frame(pcr=no.pcr, pstar = pstar, se=pstar.se)
```

Calculate 95% confidence intervals:
```{r}
pstar.df$lower <- pstar.df$pstar - pstar.df$se*1.96
pstar.df$upper <- pstar.df$pstar + pstar.df$se*1.96
pstar.df$upper[pstar.df$upper > 1] <- 1
write.csv(pstar.df, "./pstar_cq40.csv")
pstar.df
```

```{r, include=FALSE}
# Pstar plot
pstar.95 <- 5 # number of samples needed for detection to exceed 95%
pstar.80 <- 3 #80% 
pstar.50 <- 1 #50%

pstar.plot <- ggplot(data=pstar.df, aes(x=pcr, y=pstar))+
  geom_ribbon(aes(x=pcr,ymin=`lower`, ymax=`upper`),fill="gray", alpha=.6)+
  geom_line(linewidth=1)+
  geom_vline(xintercept=pstar.80, color='gray10', linetype="dashed", linewidth=0.8)+
  ylab("Cumulative detection probability (p*)")+
  xlab("Number of qPCR replicates")+
  theme_classic()+
  theme(text = element_text(size = 20)) +
  coord_cartesian(ylim = c(0.1,1), xlim = c(1,10)) +
  annotate("text", x = Inf, y = Inf, label = paste("B"), vjust = 24, hjust = 1.5, size = 8)
pstar.plot
```

Calculate theta*:
```{r}
no.samples <- seq(1,10,1) # Number of water sample replicates

tstar <- 1-((1-eDNA.out40$mean$mean.theta)^no.samples)

sigma.t <- matrix(c(eDNA.out40$sd$mean.theta^2,0, 
                  0,0), nrow=2)

tstar.se <- vector(length=length(tstar))

for (i in 1:length(no.samples)){
    tstar.se[i]<- sqrt(deltavar(1-((1-x)^y),
                       meanval = c(x=eDNA.out40$mean$mean.theta, y=no.samples[i]),
                       Sigma=sigma.t)) }

tstar.df <- data.frame(samples=no.samples, tstar = tstar, se=tstar.se)
```

95% confidence intervals:
```{r}
tstar.df$lower <- tstar.df$tstar - tstar.df$se*1.96
tstar.df$upper <- tstar.df$tstar + tstar.df$se*1.96
tstar.df$upper[tstar.df$upper > 1] <- 1
write.csv(tstar.df, "./thetastar40.csv")
tstar.df
```

```{r, include=FALSE}
# Theta* plot:
tstar.95 <- 4 # number of samples needed for detection to exceed 95%
tstar.80 <- 3 #80% 
tstar.50 <- 1 #50%

tstar.plot<- ggplot(data=tstar.df, aes(x=samples, y=tstar))+
  geom_ribbon(aes(x=samples,ymin=`lower`, ymax=`upper`),fill="gray", alpha=.6)+
  geom_line(linewidth=1)+
  geom_vline(xintercept=tstar.80, color='gray10', linetype="dashed", linewidth=0.8)+
  ylab(expression("Cumulative availability probability (" ~ theta ~ "*)")) +
  xlab("Number of 1L samples")+
  theme_classic()+
  theme(text = element_text(size = 20)) +
  coord_cartesian(ylim = c(0.1,1), xlim = c(1,10)) +
  annotate("text", x = Inf, y = Inf, label = paste("A"), vjust = 24, hjust = 1.5, size = 8)
tstar.plot
```

```{r, include=FALSE}
# Save plots
png(filename="./pstar_thetastar_cq40.png", width = 1000)
grid.arrange(pstar.plot, tstar.plot, nrow = 1)
dev.off()
```


# Presence threshold: Cq = 50

Import eDNA data (split by qPCR replicate). Most samples have 4 replicates but some have 8:
```{r}
pcr1 <- read.csv("./qpcr_rep1_cq50.csv", header=TRUE, row.names = 1) 
pcr2 <- read.csv("./qpcr_rep2_cq50.csv", header=TRUE, row.names = 1)  
pcr3 <- read.csv("./qpcr_rep3_cq50.csv", header=TRUE, row.names = 1)  
pcr4 <- read.csv("./qpcr_rep4_cq50.csv", header=TRUE, row.names = 1) 
pcr5 <- read.csv("./qpcr_rep5_cq50.csv", header=TRUE, row.names = 1) 
pcr6 <- read.csv("./qpcr_rep6_cq50.csv", header=TRUE, row.names = 1)  
pcr7 <- read.csv("./qpcr_rep7_cq50.csv", header=TRUE, row.names = 1)  
pcr8 <- read.csv("./qpcr_rep8_cq50.csv", header=TRUE, row.names = 1) 
```

Create empty array of 8, 93x5 matrices:
```{r}
A <- array(as.numeric(NA), dim = c(93,5,8))
```

Fill array with eDNA data. Array dimensions are [1:93 (sites), 1:5 (sample replicates), 1:8 (qPCR replicates)]:
```{r}
A[,,1]<-as.matrix(pcr1)
A[,,2]<-as.matrix(pcr2) 
A[,,3]<-as.matrix(pcr3) 
A[,,4]<-as.matrix(pcr4) 
A[,,5]<-as.matrix(pcr5)
A[,,6]<-as.matrix(pcr6) 
A[,,7]<-as.matrix(pcr7) 
A[,,8]<-as.matrix(pcr8) 

y <- A 
```

## WinBUGS model

Format data for WinBUGS:
```{r}
str( win.data50 <- list(y = y, # data used to fill array
                      n.site = dim(y)[1], # number of rows = number of sites (93)
                      n.samples = dim(y)[2], # number of columns = number of 1L replicate samples/site (up to 5)
                      n.pcr = dim(y)[3] )) # number of PCR replicates (up to 8)
```

Define model in BUGS language:
```{r, include=TRUE, eval=FALSE}
sink("eDNA.model50.txt") 

cat("
model{
# Priors and model for params 
  int.psi ~ dunif(0,1) # Intercept of occupancy probability (sites) 
  for(t in 1:n.samples){ 
      int.theta[t] ~ dunif(0,1)} # Intercepts of availability probability per sample replicate
                             
  for(t in 1:n.pcr){ 
      int.p[t] ~ dunif(0,1)}  # Intercepts of detection probability (1-PCR error)
      
# 'Likelihood' (or basic model structure)
  
  ## Occurrence in site i 
  for(i in 1:n.site){ 
      z[i] ~ dbern(psi[i]) 
      logit(psi[i]) <- logit(int.psi) 
      
      ## Occurrence in sample j 
      for(j in 1:n.samples){ 
          a[i,j] ~ dbern(mu.a[i,j])
          mu.a[i,j] <- z[i] * theta[i,j]
          logit(theta[i,j]) <- logit(int.theta[j])
  
          ## PCR detection error process in sample k 
          for (k in 1:n.pcr){ 
               y[i,j,k] ~ dbern(mu.y[i,j,k]) 
               mu.y[i,j,k] <- a[i,j] * p[i,j,k] 
               logit(p[i,j,k]) <- logit(int.p[k])
                            }
                          }
                tmp[i] <- step(sum(a[i,])-0.1) 
                        }

# Derived quantities 
sum.z <- sum(z[]) # Total number of occupied sites 
sum.a <- sum(tmp[]) # Total number of samples with presence 
mean.p <- mean(int.p[]) # mean p across qPCR replicates 
mean.theta <- mean(int.theta[]) # mean theta across sample replicates 

} # end model 
",fill=TRUE)

sink()
```

Initial values:
```{r}
zst <- apply(y, 1, max) # inits for presence (z) 
ast <- apply(y, c(1,2), max) # inits for availability (a): applies the "max" function to all matrix rows and columns in the array
inits <- function() list(z = zst, a = ast, int.psi = 0.5)
```

Parameters:
```{r}
params <- c("int.psi", "int.theta", "int.p", "sum.z", "sum.a", "mean.p","mean.theta", "z") 
```

MCMC settings: 
```{r}
ni <- 25000 ; nt <- 10 ; nb <- 2000 ; nc <- 3
# ni <- 50000 ; nt <- 5 ; nb <- 2000 ; nc <- 3
```

Call WinBUGS and summarize posterior:
```{r, include=TRUE, eval=FALSE}
bd <- "C:/Program Files/winbugs14_full_patched/WinBUGS14" # Location of WinBUGS

eDNA.out50 <- bugs(win.data50, inits, params, "eDNA.model50.txt", 
                 n.chains = nc, n.thin = nt, 
                 n.iter = ni, n.burnin = nb, 
                 debug = TRUE, bugs.seed = 42,
                 bugs.dir = bd)
saveRDS(eDNA.out50, "./eDNA_out50.RDS")
```

View posterior summary:
```{r, cache=TRUE}
eDNA.out50 <- readRDS("./eDNA_out50.RDS")
#print(eDNA.out40, 4)
print(eDNA.out50, 4)
```

## Calculate p-star and theta-star (code from Erik Blomberg)

Calculate p*:
```{r}
no.pcr <- seq(1,10,1) # 1-10 qPCR replicates

pstar50 <- 1-((1-eDNA.out50$mean$mean.p)^no.pcr)

sigma50 <- matrix(c(eDNA.out50$sd$mean.p^2,0, 
                  0,0), nrow=2)

pstar.se50 <- vector(length=length(pstar50)) 

for (i in 1:length(no.pcr)){
  pstar.se50[i]<- sqrt(deltavar(1-((1-x)^y),
                              meanval = c(x=eDNA.out50$mean$mean.p, y=no.pcr[i]),
                              Sigma=sigma50)) }

pstar.df50 <- data.frame(pcr=no.pcr, pstar = pstar50, se=pstar.se50)
```

95% Confidence Intervals:
```{r}
pstar.df50$lower <- pstar.df50$pstar - pstar.df50$se*1.96
pstar.df50$upper <- pstar.df50$pstar + pstar.df50$se*1.96
pstar.df50$upper[pstar.df50$upper > 1] <- 1
write.csv(pstar.df50, "./pstar_cq50.csv")
pstar.df50
```

```{r, include=FALSE}
# Pstar plot
pstar.95 <- 5 # number of samples needed for detection to exceed 95%
pstar.80 <- 3 #80% 
pstar.50 <- 1 #50%

pstar.plot50 <- ggplot(data=pstar.df50, aes(x=pcr, y=pstar))+
  geom_ribbon(aes(x=pcr,ymin=`lower`, ymax=`upper`),fill="gray", alpha=.6)+
  geom_line(linewidth=1)+
  geom_vline(xintercept=pstar.80, color='gray10', linetype="dashed", linewidth=0.8)+
  ylab("Cumulative detection probability (p*)")+
  xlab("Number of qPCR replicates")+
  theme_classic()+
  theme(text = element_text(size = 20)) +
  coord_cartesian(ylim = c(0.1,1), xlim = c(1,10)) +
  annotate("text", x = Inf, y = Inf, label = paste("B"), vjust = 24, hjust = 1.5, size = 8)
pstar.plot50
```

Calculate theta*:
```{r}
no.samples <- seq(1,10,1) # Number of water samples

tstar50 <- 1-((1-eDNA.out50$mean$mean.theta)^no.samples)

sigma.t50 <- matrix(c(eDNA.out50$sd$mean.theta^2,0, 
                  0,0), nrow=2)

tstar.se50 <- vector(length=length(tstar50))

for (i in 1:length(no.samples)){
    tstar.se50[i]<- sqrt(deltavar(1-((1-x)^y),
                       meanval = c(x=eDNA.out50$mean$mean.theta, y=no.samples[i]),
                       Sigma=sigma.t50)) }

tstar.df50 <- data.frame(samples=no.samples, tstar = tstar50, se=tstar.se50)
```

95% confidence intervals:
```{r}
tstar.df50$lower <- tstar.df50$tstar - tstar.df50$se*1.96
tstar.df50$upper <- tstar.df50$tstar + tstar.df50$se*1.96
tstar.df50$upper[tstar.df50$upper > 1] <- 1
write.csv(tstar.df50, "./thetastar50.csv")
tstar.df50
```

```{r, include=FALSE}
# Theta* plot
tstar.95 <- 4 # number of samples needed for detection to exceed 95%
tstar.80 <- 2 #80% 
tstar.50 <- 1 #50%

tstar.plot50<- ggplot(data=tstar.df50, aes(x=samples, y=tstar))+
  geom_ribbon(aes(x=samples,ymin=`lower`, ymax=`upper`),fill="gray", alpha=.6)+
  geom_line(linewidth=1)+
  geom_vline(xintercept=tstar.80, color='gray10', linetype="dashed", linewidth=0.8)+
  ylab(expression("Cumulative availability probability (" ~ theta ~ "*)")) +
  xlab("Number of 1-L eDNA samples")+
  theme_classic()+
  theme(text = element_text(size = 20)) +
  coord_cartesian(ylim = c(0.1,1), xlim = c(1,10)) +
  annotate("text", x = Inf, y = Inf, label = paste("A"), vjust = 24, hjust = 1.5, size = 8)
tstar.plot50
```

```{r, include=FALSE}
# Save plots
#png(filename="./pstar_thetastar_cq50.png", width = 1200, height = 500)
ggsave <- grid.arrange(tstar.plot50, pstar.plot50, nrow = 1)
ggsave(filename = "./pstar_thetastar_cq50.png", 
       plot = ggsave, width = 12.6, height = 6.3, dpi = 600)
#dev.off()
```

